---
title: "GSE89567_batchcorrected"
output: html_notebook
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressMessages(library(nichenetr))
suppressMessages(library(dplyr))
suppressMessages(library(tidyverse))
suppressMessages(library(Seurat))
suppressMessages(library(ggplot2))
```

# Load prior models & matrices; load datasets
### Load prior models
```{r}
lr_network <- readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
weighted_networks <- readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
# ligand_target_matrix: target genes in rows, ligands in columns
ligand_target_matrix <- readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
```

```{r}
head(lr_network)
```

```{r}
head(weighted_networks)
```

```{r}
ligand_target_matrix[1:5, 1:5]
```

```{r}
prior_models <-
  list(
    lr_network = lr_network,
    weighted_networks = weighted_networks,
    ligand_target_matrix = ligand_target_matrix
  )
```

### Load the dataset and top DEGs
```{r}
expr_genes <- read.csv("./data/GSE89567_batchcorrected_expressed_genes.csv", header = 1)
marker_genes <- read.csv("./data/GSE89567_batchcorrected_cluster_marker_genes.csv", header = 1)
```

```{r}
expr_genes[1:ncol(expr_genes)]
```
```{r}
marker_genes[1:ncol(marker_genes)]
```

### Load count matrix (filtered, global normalized and batch-corrected)
```{r}
# counts <-
#   read.csv(
#     "./data/GSE89567_batchcorrected_normalized_with_clusterID.csv.gz",
#     row.names = "Cell_ID",
#     header = 1
#   )
```

```{r}
# saveRDS(object = counts,
#         file = "./data/GSE89567_batchcorrected_normalized_with_clusterID.rds")
```

```{r}
counts <- readRDS(file = "./data/GSE89567_batchcorrected_normalized_with_clusterID.rds")
```

```{r}
# Don't run this after UPDATE counts
cluster_assignment <- counts$Cluster_ID
cluster_assignment[1:5]
```

```{r}
# UPDATE counts
counts$Cluster_ID <- NULL
counts[1:5, 1:5]
```

### Create Seurat object for the count matrix
```{r}
counts_s <- CreateSeuratObject(counts = t(counts), 
                               project = "raw")
counts_s <- AddMetaData(object = counts_s,
                        metadata = cluster_assignment,
                        col.name = "cluster.ident")
```

# Cell Communication
### Import utils
```{r}
require(gridExtra)
suppressMessages(library(hash))
source("utils.R")
```

### Functions
```{r}
runNicheNet <- function(sender, receiver, counts_s, prior_models, fig_path = "/plots/") {
  
  #' Wrapper function to run a single NicheNet analysis (Sender cluster -> Receiver cluster)
  #' @param counts_s Gene by Cell count matrix in Seurat object
  #' @param sender Name for sender cluster
  #' @param receiver Name for receiver cluster
  #' @param prior_models NicheNet prior networks
  #' @return list of prediction raw results & plots
  
  print(paste("Predicting RL interactions from", sender, "-->", receiver, sep = " "))
  
  sc_expr_genes <- expr_genes[[sender]]
  rc_expr_genes <- expr_genes[[receiver]]
  rc_marker_gsoi <- marker_genes[[receiver]] # genes of inteset (DEGs / marker genes) for the receiver cluster
  
  result_list <- get_RL_pred(count_matrix = counts_s,
                             geneset_oi = rc_marker_gsoi,
                             es = sc_expr_genes,
                             er = rc_expr_genes,
                             from_count = FALSE,
                             N = 25,
                             prior_models = prior_models,
                             name_l = sender,
                             name_r = receiver,
                             fig_path = fig_path)
  
  # visualization
  p1 <- result_list$p_ligand_receptor_network
  p2 <- DotPlot(counts_s,
                features = result_list$best_upstream_ligands %>% rev(),
                cols = "RdBu",   # "RdBu" is the palette name
                group.by = "cluster.ident"
  ) + RotatedAxis() + xlab(paste("Ligand from", sender))
  p3 <- DotPlot(counts_s,
                features = result_list$top_receptors %>% rev(),
                cols = "RdBu",
                group.by = "cluster.ident"
  ) + RotatedAxis() + xlab(paste("Receptors in", receiver)) + theme(
    axis.text.x=element_text(angle=45,size = rel(0.5), margin = margin(0.2, unit = "cm"), vjust = 1))
  
  res <- list(pred_lr_network=result_list$pred_lr_network,
              top_ligands=result_list$best_upstream_ligands, 
              top_receptors=result_list$top_receptors, 
              fig=list(lr_network=p1, dot_ligands=p2, dot_receptors=p3))
  
  return (res)
}
```


```{r}
# Wrapper function to save plots
saveNicheNetPlots <- function(fig_list, sender, receiver, fig_path = "/plots/") {
  fig_path = paste0(getwd(), fig_path)
  if (!dir.exists(fig_path)) {
    dir.create(fig_path)
  }

  tiff(filename = paste0(fig_path, "rl_", sender, '_', receiver, ".tiff"), width = 10, height = 6, units = "in", res = 300)
  print(fig_list$lr_network)
  dev.off()
  tiff(filename = paste0(fig_path, "ligand_", sender, '_', receiver, ".tiff"), width = 8, height = 5, units = "in", res = 300)
  print(fig_list$dot_ligands)
  dev.off()

  tiff(filename = paste0(fig_path, "receptor_", sender, '_', receiver, ".tiff"), width = 8, height = 5, units = "in", res = 300)
  print(fig_list$dot_receptors)
  dev.off()
}
```


### Visualization
```{r}
# # This takes a long time and generate two figures for each direction of each pair
# # but running this is necessary because it also generates 
# tumor_clusters <- c("Transformed1", "Transformed2")
# macrophage_clusters <- c("Macrophage1", "Macrophage2")
# pred_lr_network_list = list()
# for (tumor in tumor_clusters) {
#   for (macrophage in macrophage_clusters) {
# 
#     # Ligand --> Receptor: Tumor cluster --> Macrophage
#     res_tumor_to_macrophage <- runNicheNet(
#       counts_s = counts_s,
#       sender = tumor,
#       receiver = macrophage,
#       prior_models = prior_models
#     )
#     pred_lr_network_list[[paste(tumor, macrophage, sep = '_')]] <- res_tumor_to_macrophage$pred_lr_network
#     # saveNicheNetPlots(res_tumor_to_macrophage$fig, sender = tumor, receiver = macrophage)
# 
#     # Ligand --> Receptor: Macrophage --> Tumor cluster
#     res_macrophage_to_tumor <- runNicheNet(
#       counts_s = counts_s,
#       sender = macrophage,
#       receiver = tumor,
#       prior_models = prior_models
#     )
#     pred_lr_network_list[[paste(macrophage, tumor, sep = '_')]] <- res_macrophage_to_tumor$pred_lr_network
#     # saveNicheNetPlots(res_macrophage_to_tumor$fig, sender = macrophage, receiver = tumor)
#     gc()
# 
#   }
# }
```


# Calculate top Reception-Ligand interactions predicted from each tumor-macrophage cluster
```{r}
# Helper function
topPredRLscores <- function(pred_lr_network, n = 50) {
    top_prior_scores <- rev(sort(unlist(pred_lr_network)))
    top_prior_scores <- top_prior_scores[top_prior_scores > 0][1:n]
    indices <- lapply(top_prior_scores, function(x) which(pred_lr_network == x, arr.ind = TRUE))
    
    top_score_df <- data.frame(Ligand = character(), Receptor = character(), Score = numeric())
    for (i in 1:n) {
        top_score_df <- rbind(top_score_df, data.frame(
            Ligand = colnames(pred_lr_network)[indices[[i]][2]],
            Receptor = rownames(pred_lr_network)[indices[[i]][1]],
            Score = top_prior_scores[i]
        ))
    }
    
    return(top_score_df)
}
```


### Transformed 1 --> Macrophage 1
```{r}
topPredRLscores(pred_lr_network_list$Transformed1_Macrophage1)
```

### Macrophage 1 --> Transformed 1
```{r}
topPredRLscores(pred_lr_network_list$Macrophage1_Transformed1)
```

### Transformed 2 --> Macrophage 1
```{r}
topPredRLscores(pred_lr_network_list$Transformed2_Macrophage1)
```

### Macrophage 1 --> Transformed 2
```{r}
topPredRLscores(pred_lr_network_list$Macrophage2_Transformed1)
```

### Transformed 1 --> Macrophage 2
```{r}
topPredRLscores(pred_lr_network_list$Transformed1_Macrophage2)
```

### Macrophage 2 --> Transformed 1
```{r}
topPredRLscores(pred_lr_network_list$Macrophage2_Transformed1)
```

### Transformed 2 --> Macrophage 2
```{r}
topPredRLscores(pred_lr_network_list$Transformed2_Macrophage2)
```

### Macrophage 2 --> Transformed 2
```{r}
topPredRLscores(pred_lr_network_list$Macrophage2_Transformed2)
```

### Save top 50 R-L pairs to csv
```{r}
output_path <- "./top_rl_pairs/"
for (tumor in tumor_clusters) {
  for (macrophage in macrophage_clusters) {
    keyword1 <- paste(tumor, macrophage, sep = "_")
    keyword2 <- paste(macrophage, tumor, sep = "_")

    write.table(data.frame(topPredRLscores(pred_lr_network_list[[keyword1]])),
                paste0(output_path, keyword1, ".csv"), sep=',' )
    write.table(data.frame(topPredRLscores(pred_lr_network_list[[keyword2]])),
                paste0(output_path, keyword2, ".csv"), sep=',' )
  }
}
```